Permute-match tests: Detecting significant correlations between time series despite nonstationarity and limited replicates

Version 1

Abstract

Abstract

Researchers in fields from ecology to neuroscience analyze correlations between pairs of time series, often working with nonstationary data, wherein statistical properties change over time. This commonly involves a statistical test to determine whether an observed correlation is stronger than expected under the null hypothesis of independence. Testing for dependence between nonstationary time series with only one experimental replicate is exceedingly challenging. However, with many replicates, a nonparametric trial-swapping permutation test can be employed, comparing within-replicate correlations to between-replicate correlations. Although largely assumption-free, this test is severely limited by the number of replicates because its minimum achievable p-value is 1/n! where n is the number of replicates. This curtails its applicability to many biomedical studies, where n is frequently as low as 3, which would render significance thresholds like 0.05 unattainable. To address this, we propose modified permutation tests that can report lower p-values of 2/nn or 1/nn when there is strong evidence of dependence. We prove that the tests guarantee a false positive rate at or below the significance level, as long as replicates come from independent and identical experiments. We demonstrate this approach by confirming the observation that groups of zebrafish swim faster when directionally aligned, using an existing dataset with 3 biological replicates.

Introduction

Scientists frequently look for correlations between variables to identify potentially important relationships, or to support conceptual models. In disciplines with a focus on dynamics such as ecology and physiology, it is common to measure correlations between time series. Many important biological processes are nonstationary, meaning that their statistical properties (e.g. mean or variance) change systematically across time. Such processes range from the spread of an invasive species to a cell’s response to a new environmental stress.

Interpreting a correlation between a pair of nonstationary time series can be highly fraught because it is easy to obtain a seemingly impressive correlation between two time series that have no meaningful relationship [1]. For example, the sizes of any two exponentially-growing populations will be correlated over time due to a shared growth law, even if the populations lack any interaction or shared influences. To avoid overinterpreting spurious correlations, it helps to distinguish between the concepts of “correlation” and “dependence”. In time series research, the term “correlation” is often used procedurally [2, 3, 4]. Similarly, here we define a correlation function (ρ) to be any function that takes two time series and produces a statistic, although it is usually interpreted as a measure of similarity or relatedness.

In contrast to correlation, “dependence” is a hypothesis about the relationship between variables, and it has clearer scientific implications. Two events A and B are called dependent if the probability that they both occur P (A, B) differs from the product of their individual probabilities P (A)P (B). Similarly, two temporal processes are dependent if the probability of observing any particular pair of trajectories differs from the product of the probabilities of individual trajectories. (The formal definition of dependence extends this idea to continuous variables, where discrete probabilities may be replaced by probability densities [5, 6].) The importance of dependence relationships can be attributed to Reichenbach’s common cause principle, which states that if two variables are dependent, then either they share a common causal influence, or one variable causally influences the other (possibly indirectly) [7, 8]. Thus, it is useful to to first test whether an observed correlation indicates dependence before pursuing specific mechanistic explanations.

There are a handful of systematic approaches to testing for dependence between nonstationary time series. However, most are fairly limited in their scope. First, when possible, a nonstationary time series can be transformed to become stationary, meaning that its statistical properties do not change systematically across time. This enables access to a wide arsenal of tests applicable to stationary data. Transformations include subtracting a trend, taking the derivative (more precisely, “differencing” between neighboring points), or choosing a stationary-seeming window of time [9, 10]. However, it is easy to see potential pathologies in each of these: Taking the derivative of an exponential curve just produces another exponential curve; subtracting a fitted linear trend from the random walk X(t) = X(t − 1) + ϵ(t) (where ϵ(t) is random noise) does not make it stationary [9]; similarly, searching for a stationary window of the same random walk process is futile since the variance of X(t) increases with each step. In a second approach, one may compare the observed correlation with correlations obtained when one time series is replaced by a synthetic replicate, where synthetic replicates are generated in a way that reflects the original nonstationary process [11]. However, these require a correct model of how the statistical properties of the process evolve over time. Lastly, there are tests within the econometrics literature that can provide evidence for dependence between random-walk-like nonstationary processes by detecting a property called cointegration, but cointegration only occurs when some linear combination of the time series is stationary [12].

Alternatively, an approach based on permutations is possible when there are multiple identically distributed and independent (iid) replicates (or trials; we use “replicates” and “trials” interchangeably). In this case, one may evaluate the significance of a within-replicate correlation by comparing it to between-replicate correlations. That is, if Xi and Yi are time series of variables X and Y from replicate i, the correlation of (Xi, Yi) may be compared to the correlation of (Xi, Yj≠ i). If the two variables are dependent, within-replicate correlations should tend to be stronger than between-replicate correlations. This approach is sometimes called “inter-subject surrogates” [13, 14].

The inter-subject surrogate approach can be used to test for dependence in each trial separately [15]. In this case, a simple nonparametric test of dependence between, for instance, X1 and Y1 can be performed by computing the correlation ρ(X1, Yj) for j = 1, …, n and writing down a p-value as the proportion of these correlations that are greater than or equal to ρ(X1, Y1) [16, 17]. For this approach, at least n = 20 trials are needed if one wishes to possibly obtain a p-value of 0.05 or below. This procedure checks for dependence in trial 1 specifically; assessing dependence in a different trial via this method would involve an analogous test.

To reduce the required number of trials below 20, a straightforward strategy is to perform a single permutation test (Fig 1A) using the data from all replicates [18, 19]. As an example with n = 4 trials, the test procedure begins by computing the mean within-trial correlation:

Next, as a null model, recompute the mean correlation after permuting the Y time series while holding the X time series in the original order. This is the same as permuting Xs and maintaining the order of the Y s. For instance, one such permutation might be:

A p-value can be calculated as the proportion of permutations (including the original ordering) that produce a mean correlation that is as strong as, or stronger than, ρorig (Fig 1A). One may then detect a correlation at the significance level α if p ≤ α. We emphasize that these permutations are obtained by swapping trials, not by swapping time points. This test has been used to detect correlations between time series in neuroscience and psychology settings [20, 18, 21]. It has also been used to detect correlations between variables measured in brain images, which can have similar nonstationarity challenges as time series [22]. A noteworthy advantage of this approach is that it is valid (meaning that the false positive rate is guaranteed to not exceed α) under very mild assumptions. Namely, the test is valid if the Xi trials are exchangeable with one another (i.e. all permutations of the sequence X1, …, Xn have the same joint probability distribution), or if the Yi trials are exchangeable (see Corollary 4 in Appendix 1).

The permutation test, perfect match concept, and permute-match tests. (A) The permutation test. Circles indicate the average correlation obtained with (grey) or without (pink) trial swapping. Here, N≥ is the number of average correlations (including the original correlation) that are equal to or larger than the original correlation. pperm is then N≥/n! because n! is the total number of average correlations, including the original. (B) Illustration of a Y -perfect match, represented as a table of correlations wherein each diagonal entry is the greatest in its own row. Each line with a “>” or “<” symbol denotes a required ordering relationship between the two numbers at either end of the line. Note that this example has a Y -perfect match, but not an X-perfect match. For an X-perfect match, each diagonal entry would need to be the greatest in its own column. (C) If an X-or Y -perfect match occurs, then the original correlation, ρorig is always greater than any shuffled correlation ρshfl. Top: If a Y -perfect match occurs, then each Yi gives the greatest correlation when it s paired with the Xi from the same trial. A shuffled correlation ρshfl pairs Yis from some trials with Xjs from different trials, thereby reducing the correlation. Lines with symbols “>“, “<“, and “=” denote comparisons between terms. Bottom: If an X-perfect match occurs, a similar argument can be applied. (D) Since a perfect match ensures that the original correlation is greater than any shuffled correlation, a perfect match also ensures that pperm takes on its lowest possible value of 1/n!. (E) The simultaneous permute-match(X;Y) test. (F) The sequential permute-match(X▸Y) test; note that the permute-match(Y ▸X) test is defined analogously.

Yet, even the permutation test remains considerably limited by the number of trials. The lowest p-value that can be obtained with n replicates is 1/n! since there are n! possible permutations. For example, with n = 3, the lowest possible p-value is 1/3! ≈ 0.17 and with n = 4, the lowest is 1/4! ≈ 0.04. These minima apply even if the average within-trial correlation is much greater than the average between-trial correlation. While replicate counts of 4 and 5 can produce p-values below the conventional 0.05 threshold, these p-values remain modest. This limitation is particularly problematic when multiple hypothesis testing is required, as the modest p-values may no longer be significant after applying corrections such as the Bonferroni method.

The challenge of limited replicates is a practical reality in many experimental settings due to cost and feasibility concerns [23, 24, 25]. Factors outside the control of investigators further restrict the number of useful replicates. For example, a recently published study of cerebral ischemia in pigs began with 6 animals, but lost 2 due to death and experimental mishap, so complete records exist for only 4 pigs [26]. Similarly, in an animal migration study, authors affixed GPS tags to 13 juvenile cuckoos, but only 5 of them produced GPS records suitable for the study due to heterogeneous behavior among cuckoos [27]. Finally, secondary analysis of public data from earlier studies can be a resource-efficient mode of research [28], and this approach is necessarily limited to the number of replicates within the existing dataset.

In this article, we show that by adding additional steps to the permutation test, we may achieve p-values as low as 1/nn. For n = 3 or 4 respectively, the lowest possible p-value is then ≈ 0.04 or 0.004. This new result only requires that the replicates be iid. Thus, this modified permutation test allows the data analyst to detect dependence with stronger confidence when there is sufficient evidence to do so. We illustrate this approach using simulations as well as a real world example involving recordings of zebrafish swimming behavior with only 3 replicates.

Results

The concept of an X-perfect or Y -perfect match

Central to our approach is the concept of an X-perfect match or Y -perfect match. To motivate the ideas, suppose that a group of n graduate students X = (X1, …, Xn) has been paired with a group of n advisors Y = (Y1, …, Yn) so that the ith student (Xi) is paired with the ith advisor (Yi). Moreover, let ρ(Xi, Yj) be a number that measures how well the ith student and the jth advisor get along. We say that the ith student is “happy” if they get along with their own advisor strictly better than they get along with any other advisor, meaning that ρ(Xi, Yi) > ρ(Xi, Yj) for all j ≠ i. Similarly, the ith advisor is “happy” if ρ(Xi, Yi) > ρ(Xj, Yi) for all j ≠ i. The arrangement of student-advisor pairs is X-perfect if all students are happy, and Y -perfect if all advisors are happy.

Analogously, if X = (X1, …, Xn) and Y = (Y1, …, Yn) are collections of time series with n trials each, and if ρ is some correlation function, we say that an X-perfect match has occurred if (and only if) for all X trials,

Similarly, a Y -perfect match has occurred if and only if:

Throughout the article, we use boldface X or Y to indicate a collection of trials, Xi or Yi to indicate an individual trial, and X or Y to refer to the variable in general.

A perfect match (either X-perfect or Y -perfect) provides especially strong evidence to support the hypothesis that X and Y are dependent. With a perfect match, the original correlation ρorig is strictly greater than any of the shuffled correlations, as illustrated in Fig 1C. Thus, a perfect match guarantees pperm = 1/n!, the lowest p-value achievable with the permutation test (Fig 1D). Furthermore, letting P denote probability, we can prove that under the null hypothesis H0 wherein (1) the Yi trials are iid, (2) the Xi trials are iid, and (3) X and Y are independent (Lemma 6 in Appendix 2). By an analogous argument (Lemma 7 in Appendix 2), under H0 we have P (Y −perfect match) ≤ 1/nn and thus

Permute-match tests: modified permutation tests of dependence

We now describe tests for dependence between nonstationary time series based on the ideas just laid out, which we will call permute-match tests. Suppose that an experiment produces time series X = (X1, …, Xn) and Y = (Y1, …, Yn), where Xi and Yi are time series obtained from the ith replicate, and we want to test whether X and Y are dependent. For example, we may have video recordings of n animals in separate but identically constructed enclosures; for the ith animal, we extract a time series of movement speed (Xi) and a time series of the distance from a light source (Yi). Using these data, we may wish to investigate whether the animals under study tend to move at different speeds depending on their distance from the light source. We introduce two tests: simultaneous permute-match and sequential permute-match. As we will see, the former can be more reproducible while the latter can be more powerful.

In the simultaneous permute-match test (Fig 1E), we check for both an X−perfect and a Y − perfect match. If either the X-or Y -perfect match occurs, we write down a p-value to be p = 2/nn per equation 1. If neither X nor Y achieves a perfect match, then we fall back to the permutation test: p = pperm = N≥/n! where N≥ is the number of all n! possible permutations (including the original ordering) that produce an average correlation that is at least as large as the original ordering (Fig 1A).

In the sequential permute-match test, we begin by choosing a variable (either X or Y) for the first perfect match test. For example, suppose we choose X first (in which case we refer to the procedure as the “permute-match(X▸Y)” test). If an X-perfect match is observed, then p = 1/nn (Fig 1F). If X-perfect match is not observed, we then check for a Y -perfect match, and write p = 2/nn if a Y -perfect match is observed. If neither an X-perfect nor Y -perfect match occurs, then p = pperm. A permute-match(Y ▸X) test is defined analogously.

Under H0 (i.e. the Yi replicates are iid, the Xi replicates are iid, and X and Y are independent), all the above tests are valid, meaning that for any significance level α, we have P (p ≤ α) ≤ α. This fact is proven as Theorems 10-12 in Appendix 2.

The permute-match tests (Fig 1E-F) are “upgrades” to the permutation test (Fig 1A) because permute-match tests always report the same or a lower p-value (since 1/nn ≤ 2/nn < 1/n! for all integers n ≥ 2; see Lemma 9 in Appendix 1). In exchange for lower p-values, the permute-match tests have a slightly more restrictive data requirement than the permutation test. The permutation test is valid either if the Xi trials are exchangeable, or if the Yi trials are exchangeable (Corollary 4 in Appendix 1). The permute-match tests require that all Xi trials are iid and that Yi trials are iid. If trials are iid, then they are necessarily exchangeable. However, exchangeable trials are not always iid. For example, the first five cards drawn from a shuffled deck are exchangeable because all possible orderings of cards are equally likely, but they are not iid because if the first card is an ace of hearts, the second card cannot be the ace of hearts. In practice, biological experimental designs typically use replicates that are both exchangeable and iid, so the more restrictive requirement of the permute-match test is often inconsequential.

The different tests - permutation test, simultaneous permute-match test, and sequential permute-match tests - are valid tests of dependence given iid replicates (Theorems 10-12 in in Appendix 1). In other words, they all guarantee false positive rates at or below the significance level. Two subtle aspects of the permutematch tests are worth mentioning here. First, no multiple testing correction is applied to account for the fact that both match tests and a permutation test are used. In other words, although the simultaneous permute-match(X;Y) test assigns p = 2/nn instead of p = 1/nn when an X-or Y -perfect match occurs as a Bonferroni-type correction, no such correction is needed when combining the match test and the permutation test (Fig 1E). This is because the permutation test and perfect match events are nested. That is, an X-or Y -perfect match is possible only if pperm is already at its lowest possible value of 1/n! (Fig 1C-D). Thus, unlike combining un-nested tests where each can separately contribute false positives, we can safely combine the checks for perfect matches with the permutation test. The second subtlety refers specifically to the sequential permute-match tests. These tests use the lower p-value of 1/nn if the first perfect match occurs, but use the more conservative “Bonferroni-corrected” p-value of 2/nn if only the second match occurs (Fig 1F). This superficially resembles the practice of conducting additional tests merely because earlier tests did not result in a detection -a form of p-hacking that generally results in an invalid final p-value. However, our sequential permute-match procedure is actually valid because it combines binary events instead of continuous variables. See Appendix 3 for a detailed discussion of this point.

Permute-match test variants may differ in statistical power

Although the different tests - permutation test, simultaneous permute-match test, and sequential permutematch tests - are all valid in the sense of correctly controlling false positive rates, how do they vary in power - the probability of detecting true dependence? The answer depends on the number of replicates n and the desired significance level α. There are four regimes to consider (Fig 2A).

Statistical power of various permutation test variants as illustrated using a simple nonstationary system. (A) Summary of test power as a function of the significance level α and the number of replicates n. (B, C) System equations and example dynamics. The processes X and Y are given by a linear trend with additive noise on a time grid t = 1, 2, …, 100. The noise terms ϵX,i(t) and ϵY,i(t) are drawn from a bivariate normal distribution with a mean of 0, variance of 1, and covariance of rX,Y. The chart shows an example pair of time series where X and Y are dependent (rX,Y = 0.3). (C) Statistical power of the permutation test and various permute-match tests as a function of the replicate number n, significance level α, and strength of dependence rX,Y. Power was calculated from 5000 simulations at each value of rX,Y between rX,Y = 0 and 0.54 in steps of size 0.01. We chose the Pearson correlation coefficient as our correlation function ρ.

Regime 1: when α ≥ 1/n!, all tests have the same power. If the permutation test detects dependence, permute-match tests will too since they fall back to a permutation test in the “worst case” (Fig 2A, column 1). Conversely, if the permutation test did not detect dependence (i.e. pperm > α), then it follows that pperm > 1/n!, which implies that neither an X-perfect nor Y -perfect match occurred (Fig 1D), and so the permute-match tests cannot have detected dependence either.

Regime 2: when 1/n! > α ≥ 2/nn, all permute-match tests are tied for highest power (Fig 2A, column 2). In this regime, the permutation test has zero power, since its p-value is limited at 1/n!. Conversely, if either an X-perfect or a Y -perfect match occurs, then, the sequential permute-match tests as well as the simultaneous permute-match(X;Y) test will all detect dependence.

Regime 3: when 2/nn > α ≥ 1/nn, the simultaneous permute-match(X;Y) test will have no power as the desired p-value is below the lowest possible p-value achievable by the test. Only the permute-match(X▸Y) or (Y ▸X) tests now have power (Fig 2A, column 3). Note that permute-match(X▸Y) and permute-match(Y ▸X) test may differ in power, as can be seen in the upper left panel of Fig 2D. We do not currently know how to pre-diagnose which of these two tests will be most powerful based on a given dataset, so the choice of which test to run in this case is arbitrary.

Regime 4: when α < 1/nn, no test has any power even when a perfect match occurs.

Figure 2B-D illustrates these different scenarios using a simulated nonstationary system and some common significance cutoffs. We simulated time series of two variables (X and Y) from a small number of replicates (between 3 and 5). Here, the time series were produced from a pair of linear time trends with coupled additive noise (rX,Y determines the strength of coupling; Fig 2B). We chose the Pearson correlation coefficient as our correlation function ρ. The various sub-panels of Fig 2D illustrate the different aforementioned regimes.

The above example was chosen for ease of presentation - in fact, the example is so simple that it may be correctly treated by just fitting and removing the linear trend. In Appendix 4, we show that the same ideas hold in an example with a more complex nonstationary data-generating process (a logistic map with time-varying parameters) and a nonlinear correlation function (cross-map skill).

An example from animal behavior science

Living in groups is a common experience among animals. For example, at least half of fish species are thought to form groups at some stage of life [29, 30]. The properties of these groups can impart a variety of important fitness effects [31]. In groups of fish, coordinated swimming may facilitate foraging (e.g. by enabling fish to “communicate” the location of food to others) and predator escape (e.g. by confusing predators) [32, 30]. One study [30] used videos of small groups of zebrafish (8 fish per group) to observe that fish swam faster during video segments when they were in a high-polarization state (i.e. when they were directionally aligned) than during segments when they were in a low-polarization state. A correlation between polarization and speed might indicate statistical dependence between the two variables, or it might just be a consequence of temporal trends (e.g. polarization and swimming speed both happen to decrease with time).

Using a publicly available dataset [33] containing fish trajectories from 3 replicate 10-minute videos of small groups of juvenile Danio rerio zebrafish (10 fish per group), we performed a permute-match test of dependence between polarization and fish swimming speed. As we will see, although these quantities are potentially nonstationary, the permute-match test nevertheless detects a statistical dependence between them.

To quantify polarization, we use “circular variance” (vcirc; see Methods), a common measure of angular dispersion [34]. Since high polarization means low dispersion, we quantify polarization as 1 − vcirc, and call this quantity the circular concentration, since it indicates “how concentrated the [circular] data is toward the center” ([35], page 15). Circular concentration is bounded between 0 (low alignment) and 1 (perfect alignment). To quantify speed, we took the “average individual speed”, which is a term we use to denote an average across individuals, not across time. More precisely, the average individual speed of a 10-fish group at time t is (s1,t + s2,t + · · · + s10,t)/10 where si,t is the speed of fish i at time t. Note that average individual speed is distinct from the speed of the group center, which has also been studied in Danio fish [31].

Neither average individual speed nor circular concentration is obviously stationary (Fig 3A). By visual inspection, the average individual speed seems to decrease across time, and all time series are deemed nonstationary by a Kwiatkowski–Phillips–Schmidt–Shin test (p < 0.01), and so we may be on shaky ground to use methods that require stationarity. Additionally, since only 3 trials are available, the permutation test and the simultaneous permute-match test cannot detect dependence at the 0.05 level, so we perform a sequential permute-match test.

A sequential permute-match test detects dependence between average individual speed and circular concentration in small groups of zebrafish. (A) Time series of average individual speed and circular concentration for the three replicate videos. Black curves show original time series, and red curves show a 30-second moving average. Average individual speed appears to decrease with time in all trials. All 12 time series (2 variables × 3 trials × 2 smoothing conditions) were deemed nonstationary by a Kwiatkowski– Phillips–Schmidt–Shin (KPSS; [37]) test (p < 0.01). Note that the KPSS test seeks to reject a stationary null hypothesis in contrast to other common tests, whose null hypotheses are nonstationary. (B) Pearson correlation between circular concentration and average individual speed, both within trials and between trials, using the full 10 minutes of data. Table entries are shaded by correlation. We observe a speed-perfect match (and a circular concentration-perfect match), and thus detect dependence with p = 1/33 ≈ 0.04. (C) A parametric test also detects dependence. As a parametric alternative, for each trial we averaged values of speed and circular concentration over the full 10 minutes (e.g. the scatter plot shown, where each point is one trial). The sample Pearson correlation of the time-averaged variables is 0.99996, and a one-tailed test of significance gives p ≈ 0.003. (D) Permute-match tests detected a significant correlation between speed and circular concentration more consistently than the parametric test. We truncated time series to different lengths starting from the first frame, and compared the parametric test with the two possible permute-match tests. Length was varied between 20 seconds and 10 minutes in 20-second increments, with each increment representing 640 frames since the frame rate is 32 frames per second. We used the Pearson correlation (not its magnitude) in the permute-match tests since the alternative hypothesis is that the correlation is positive, not merely nonzero.

We arbitrarily started with average individual speed and identified a “speed”-perfect match, indicating a significant correlation between average individual speed and circular concentration (Fig 3B; p = 1/33 ≈ 0.04). Significance does not depend on the choice of which match is tested, since both a “speed”-perfect and a “circular concentration”-perfect match is obtained. Note that although trials are independent of each other, all between-trial correlations from the full dataset are positive (Fig 3B), illustrating the danger in applying naive data analysis methods to data that are autocorrelated or even nonstationary.

To try a parametric alternative, for each trial we averaged values of average individual speed and circular concentration over the full 10 minutes (Fig 3C; where each point is one trial). The sample Pearson correlation of the time-averaged variables is 0.99996, and a one-tailed test of significance (using the function stats.pearsonr from the Python package Scipy [36]) also detects dependence (p ≈ 0.003), consistent with the permute-match tests. This test relies on a null distribution derived under the assumption that the three points in Fig 3C are drawn from a bivariate Gaussian distribution with zero covariance. We view this test as comparable to the permute-match tests because although it requires a parametric assumption, it can be valid even when the time series are nonstationary.

Since data are limited in many scientific applications, we asked whether the results of the sequential permute-match tests and the parametric Pearson correlation test were consistent across different time series lengths. We truncated time series to various lengths (from 20 seconds to 10 minutes), and checked whether each test reported a significant (p ≤ 0.05) correlation (Fig 3D). Both sequential permute-match tests reported a significant correlation at all lengths tested, whereas the parametric test was inconsistent, finding a significant correlation for about one third of lengths.

Discussion

Outside the context of time series, permutation tests have been celebrated because they require only minimal assumptions [38]. To satisfy the requirements of permutation tests, it is sufficient (see Corollary 4 in Appendix 1) to collect data from exchangeable replicates. Data in each trial can be autocorrelated (e.g. temporally or spatially) and nonstationary. No distributional assumptions are required, and the correlation function can even be asymmetric so that ρ(X, Y) ≠ ρ(Y, X) as occurs when correlations are based on techniques that use one time series to predict values of another (e.g. Appendix 4). Such radical freedom from assumptions stands in stark contrast to the situation in the analysis of single-replicate time series, where the practitioner may be forced to make assumptions that can be difficult to verify. For example, most statistical tests of correlation between two time series require stationary data at a minimum [39]. Yet, it is difficult to verify that a single time series is stationary because stationarity is a property of an entire ensemble of time series [39], so that checking for stationarity in a single time series is philosophically analogous to checking for the Gaussianity of a single data point.

However, applying a trial-swapping permutation test to check for significant correlations requires that a sufficient number of trials be available to permute amongst one another. When only small numbers of trials are available, the permutation test is severely limited by mathematics. The minimum p-value that can possibly be achieved by a permutation test is 1/n!, which will provide only modest evidence against a null hypothesis when n = 4 and is essentially useless when n = 3, regardless of how strong the true dependence may be. This limitation is particularly severe for researchers testing several related hypotheses, as this ideally requires a test that can produce p-values sufficiently low to maintain significance even after a multiple-testing correction.

Here we have described modified permutation tests–the sequential and simultaneous permute-match tests–that have access to lower p-values. In the sequential permute-match test, if the first variable does not display a perfect match, this test falls back to the simultaneous permute-match test. Thus, the sequential permute-match test is more powerful than the simultaneous test. However, the sequential test requires the arbitrary choice of checking for an X-or Y -perfect match first, making it potentially less reproducible between studies than the simultaneous test, which does not involve an arbitrary choice. If neither an X-nor Y -perfect match is obtained, both tests default to a permutation test. This step is valid because the tests are nested in the sense that a perfect match can only occur when the permutation p-value takes on its lowest possible value (Lemma 8 in Appendix 1).

It seems likely that the tests could be further modified to report an even lower p-value when both an X-and Y -perfect match occur, as in Fig 3B. However, we have not yet determined a sharp upper bound on the probability of this two-way match event. This problem is left for future efforts.

Although we have here used language and examples involving time series, the permute-match test can in principle be applied to other scenarios where iid replicates are available, but data within each replicate are dependent. One example is spatial data in brain imaging, where the permutation test has been performed using multiple scanned brains [22]. Beyond spatial processes, dependent data also appear in nucleotide sequences and natural language text. Overall, advances in methods that take advantage of multiple replicates may facilitate statistical testing without requiring assumptions that are difficult to verify.

Methods

Data preprocessing

We obtained fish trajectory data [33] from the web address at https://drive.google.com/drive/folders/1UmzlX-yJhzQ5KX5rGry8wZgXvcz6HefD. For the first trial we used the file at the path “10/1/trajecto-ries_wo_gaps.npy” (where “10” indicates the number of fish and “1” indicates the first trial). For the second and third trials, the file path was the same except it began with “10/2/” and “10/3/” respectively. These files contain sequences of fish positions indexed by video frame, as well as constants such as frame rate and approximate fish body length.

We estimated the velocity of each fish as

Where is a 2-dimensional column vector specifying the position of the ith fish in units of body length at video frame is a 2-dimensional column vector specifying the fish velocity in units of body length per second at video frame t, and R is the frame rate (32 frames per second). A group’s average individual speed at each time was then given by where N is the number of fish. The circular concentration Ct of a group at each time was given by the formula where

Here, atan2 is the 2-argument arctangent function (i.e. “arctan2” in the Python package Numpy, ver. 1.21.6). Also, vx,i,t and vy,i,t denote the x and y components of the velocity vector . The calculation was carried out using the function “stats.circstats.circvar” in the Python package Astropy (ver. 3.2.2) [40].

Statistical analysis

The permute-match test is described in Fig 1 and its validity is proved in Appendix 2. For the KPSS test, we used the function “statsmodels.tsa.stattools.kpss” in the Python package Statsmodels (ver. 0.10.1) [41]. We set the “regression” argument to “c” (which tests the null hypothesis of stationarity rather than trend-stationarity), and we set the “nlags” argument to “auto” (which means that the number of lags used for testing stationarity is automatically chosen by a data-dependent method).

eLife Assessment

This manuscript reports an important new statistical method for calculating the significance of correlations between two time-series, which provides more accuracy than other methods when the data has few replicates. The proposed method solves a real-life problem that is frequently encountered and is broadly applicable to many realistic datasets in many experimental contexts. The technique is supported with compelling mathematical derivations as well as analysis of both computer-generated and previously published experimental data.

Reviewer #1 (Public review):

Summary:

The manuscript puts forward a statistical method to more accurately report the significance of correlations within data. The motivation for this study is two-fold. First, the publication of biological studies demands the report of p-values, and it is widely accepted that p-values below the arbitrary threshold of 0.05 give the authors of such studies justification to draw conclusions about their data. Second, many biological studies are limited by the number of replicate samples that are feasible, with replicates of less than 5 typical. The authors report a statistical tool that uses a permute-match approach to calculate p-values. Notably, the proposed method reduces p-values from around 0.2 to 0.04 as compared to a standard permutation test with a small sample size. The approach is clearly explained, including detailed mathematical explanations and derivations. The advantage of the approach is also demonstrated through analysis of computer-generated synthetic data with specified correlation and analysis of previously published data related to fish schooling. The authors make a clear case that this method is an improvement over the more standard approach currently used, and also demonstrate the impact of this methodology on the ability to obtain p-values that are the standard for biological research. Overall, this paper is very strong. While the subject matter seems somewhat specialized, I would make the case that this will be an important study that has broad general interest to readers. The findings are very general and applicable to many research contexts. Experimentalists also want to report accurate p-values in their work and better understand how these values are calculated. Although I believe the previous statement is true, I am not sure that many research groups doing biological work are reading specialized statistics journals regularly. Therefore a useful and broadly applicable statistical tool is well placed in this journal.

Strengths:

The proposed method is broadly applicable to many realistic datasets in many experimental contexts.

The power of this method was demonstrated with both real experimental data and "synthetic" data. The advantages of the tool are clearly reported. The zebrafish data is a great example dataset.

The method solves a real-life problem that is frequently encountered by many experimental groups in the biological sciences.

The writing of the paper is surprisingly clear, given the technical nature of the subject matter. I would not at all consider myself a statistician or mathematician, but I found the text easy to follow. The authors did an impressive job guiding the reader through material that would often be difficult to grasp. The introduction was also well-written and clearly motivated the goals of the study.

Weaknesses:

A few changes could be made if the manuscript is revised. I would consider all of these points minor, but the paper could be improved if these points were addressed.

(1) The caption of Figure 2 doesn't seem to mention panel D. Figure A-2 also does not mention C in the caption.

(2) Figure 2D is a little hard to follow. First, the definition of "Power" is not clear, and I couldn't find the precise definition in the text. Second, the legend for the different lines in 2D is only given in Figure A-2. Perhaps a portion of the caption for Figure 2 is missing?

(3) The concept of circular variance for the fish data was heard to understand/visualize. The equation on line 326 did not help much. If there is a very simple picture that could be added near line 326 that helps to explain Ct and theta, that could be a big help for some readers who do not work on related systems. The analysis performed is understandable, the reader just has to accept that circular variance captions the degree of alignment of the fish.

(4) For the data discussed in Figure 3, I wasn't 100% sure how the time windows were selected. In the caption, it says "time series to different lengths starting from the first frame". So the 20 s time window was from t=0 to t= 20 s. Would a different result be obtained if a different 20 s window was chosen (from t = 4 min to t = 4 min 20 s just to give a specific example). I suppose by chance one of the time windows would give a p-value less than the target 0.05, that wouldn't be surprising. Maybe a random time window should be selected (although I am not indicating what was reported was incorrect)? A little more discussion on this aspect of the study may be helpful.

Reviewer #2 (Public review):

Summary:

This paper presented a hypothesis testing procedure for the independence of two time-series that was potentially suitable for nonlinear dependence and for small-sample cases. This should bring potential benefits for biology data.

Strengths:

The test offers good flexibility for different kinds of dependence (through adjusting \rho), and seems to have good finite sample performance compared to the literature. The justification regarding the validity of the test procedure is clear.

Weaknesses:

(1) The size of the test is not guaranteed to (asymptotically) equal \alpha, which may damage the power.

(2) The computational time can be an issue for a moderately large sample size when calculating the X / Y-perfect match. It will be beneficial to include discussions on the implementations of the test.

Author response:

Public Reviews:

Reviewer #1 (Public review):

Summary:

The manuscript puts forward a statistical method to more accurately report the significance of correlations within data. The motivation for this study is two-fold. First, the publication of biological studies demands the report of p-values, and it is widely accepted that p-values below the arbitrary threshold of 0.05 give the authors of such studies justification to draw conclusions about their data. Second, many biological studies are limited by the number of replicate samples that are feasible, with replicates of less than 5 typical. The authors report a statistical tool that uses a permute-match approach to calculate p-values. Notably, the proposed method reduces p-values from around 0.2 to 0.04 as compared to a standard permutation test with a small sample size. The approach is clearly explained, including detailed mathematical explanations and derivations. The advantage of the approach is also demonstrated through analysis of computer-generated synthetic data with specified correlation and analysis of previously published data related to fish schooling. The authors make a clear case that this method is an improvement over the more standard approach currently used, and also demonstrate the impact of this methodology on the ability to obtain p-values that are the standard for biological research. Overall, this paper is very strong. While the subject matter seems somewhat specialized, I would make the case that this will be an important study that has broad general interest to readers. The findings are very general and applicable to many research contexts. Experimentalists also want to report accurate p-values in their work and better understand how these values are calculated. Although I believe the previous statement is true, I am not sure that many research groups doing biological work are reading specialized statistics journals regularly. Therefore a useful and broadly applicable statistical tool is well placed in this journal.

Strengths:

The proposed method is broadly applicable to many realistic datasets in many experimental contexts.

The power of this method was demonstrated with both real experimental data and "synthetic" data. The advantages of the tool are clearly reported. The zebrafish data is a great example dataset.

The method solves a real-life problem that is frequently encountered by many experimental groups in the biological sciences.

The writing of the paper is surprisingly clear, given the technical nature of the subject matter. I would not at all consider myself a statistician or mathematician, but I found the text easy to follow. The authors did an impressive job guiding the reader through material that would often be difficult to grasp. The introduction was also well-written and clearly motivated the goals of the study.

We appreciate the reviewer’s summary of our study and its strengths.

Weaknesses:

A few changes could be made if the manuscript is revised. I would consider all of these points minor, but the paper could be improved if these points were addressed.

(1) The caption of Figure 2 doesn't seem to mention panel D. Figure A-2 also does not mention C in the caption.

We apologize for this error, and thank you for catching it! The figure legends had missing or incorrect panel labels. This error has been corrected.

(2) Figure 2D is a little hard to follow. First, the definition of "Power" is not clear, and I couldn't find the precise definition in the text. Second, the legend for the different lines in 2D is only given in Figure A-2. Perhaps a portion of the caption for Figure 2 is missing?

We have added a definition of power in the main text:

“Although the permutation test, simultaneous permute-match test, and sequential permute-match test are all valid, they vary in power – the probability of detecting true dependence.”

We have clarified the use of “power” in legend of Fig 2 and clarified that the color key for Fig 2D is in Fig 2A. The relevant excerpt of the Fig 2 legend is copied here:

“(D) Statistical power for the permutation test and various permute-match tests as a function of the replicate number n, significance level α, and strength of dependence rX, Y. Power was estimated as the proportion of simulations in which dependence was detected, calculated from 5000 simulations at each value of rX, Y between rX, Y = 0 and 0.54 in steps of size 0.01. At rX, Y = 0, there is no dependence, so the curve at that point indicates the false positive rate rather than power. We chose the Pearson correlation coefficient as our correlation function ρ. See (A) for the color legend.”

We have also added dotted lines connecting the legend in panel A to the curves in panel D.

(3) The concept of circular variance for the fish data was heard to understand/visualize. The equation on line 326 did not help much. If there is a very simple picture that could be added near line 326 that helps to explain Ct and theta, that could be a big help for some readers who do not work on related systems. The analysis performed is understandable, the reader just has to accept that circular variance captions the degree of alignment of the fish.

We have replaced references to circular concentration with “mean resultant length”, which is the standard jargon for this term in circular statistics, and we have added an illustration.

(4) For the data discussed in Figure 3, I wasn’t 100% sure how the time windows were selected. In the caption, it says “time series to different lengths starting from the first frame”. So the 20 s time window was from t=0 to t= 20 s. Would a different result be obtained if a different 20 s window was chosen (from t = 4 min to t = 4 min 20 s just to give a specific example). I suppose by chance one of the time windows would give a pvalue less than the target 0.05, that wouldn’t be surprising. Maybe a random time window should be selected (although I am not indicating what was reported was incorrect)? A little more discussion on this aspect of the study may be helpful.

As suggested by the reviewer, we have redone the analysis of Figure 3D with random segments. This provides a more complete picture of how the chance of detecting a significant correlation varies with segment length. The main conclusion is unchanged: Perfect match tests reliably detect dependence across a wider range of segment lengths than the naive parametric alternative.

The relevant panel and an excerpt from the legend text are copied below.

“(D) Permute-match tests detected a significant correlation between speed and alignment more consistently than the parametric test. For a grid of lengths between 20 and 600 seconds we sampled 500 random segments of each length, each drawn from the first 600 seconds, and determined for each segment whether the parametric test and/or the two possible permute-match tests detected a significant (p ≤ 0.05) correlation. In the edge case of the maximum 600-second length, all 500 “random” segments were identical.”

Reviewer #2 (Public review):

Summary:

This paper presented a hypothesis testing procedure for the independence of two timeseries that was potentially suitable for nonlinear dependence and for small-sample cases. This should bring potential benefits for biology data.

Strengths:

The test offers good flexibility for different kinds of dependence (through adjusting \rho), and seems to have good finite sample performance compared to the literature. The justification regarding the validity of the test procedure is clear.

We appreciate the reviewer’s summary of key aspects of our manuscript.

Weaknesses:

(1) The size of the test is not guaranteed to (asymptotically) equal \alpha, which may damage the power.

We thank the reviewer for raising the issue of test size and power. We agree that a conservative test (one whose size can fall below alpha) may sacrifice power.

Our objective is distribution-free false-positive rate (FPR) control. That is, we wish to keep the FPR at or below alpha for every distribution of X and Y, because in our regime (nonstationary time series with few independent replicates) the scientist often cannot verify distributional assumptions. Inspired by the reviewer’s comment, we now show (new Proposition 14) that the perfect match probability can be made arbitrarily close to 1/n!. As a consequence, any reported perfect match p-value below 1/n! would break the distribution-free validity of the test.

A test that exploits distributional structure could likely access lower p-values; we have now explored how the empirical FPR of the permute-match test varies with the data-generating process (see our response to reviewer 2's recommendation 1 below).

(2) The computational time can be an issue for a moderately large sample size when calculating the X / Y-perfect match. It will be beneficial to include discussions on the implementations of the test.

We agree this is an important consideration. We have added the following text to the Discussion:

“The test appears computationally tractable for relevant sample sizes: Our implementation of the permute match(X>Y) procedure completed a single test of dependence in the setting of Fig 2 with an average runtime of 3 seconds when n = 10 on a 2023 14-inch MacBook Pro with an M2 Pro processor and 16 GB RAM (see Source data 1). For n > 10, a standard permutation test already can report a p-value below 3 × 10−8 so the perfect match test is likely unnecessary for typical applications.”

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

A few more minor notes/comments:

As a personal preference, I like it when figures printed in grayscale retain their meaning (when possible). Just FYI, Figure 3D in grayscale is uninterpretable. Not saying a change is needed, just pointing it out.

We changed Fig 3D to address a comment above, and think the new version better distinguishes between the permute-match and permutation test results in greyscale.

On line 304, is that a lower bound or an upper bound? Maybe the issue is the probability mentioned on line 305 is not clear.

As this point is not the main focus of the investigation, we have rephrased it to make it less technical and eliminate the issue of which bound is in question.

“It seems likely that the tests could be further modified to report an even lower p-value when an X- and Y -perfect match occur simultaneously, as in Fig 3B. However, we have not investigated further and this problem is left for future efforts.”

I am not sure if this is a weakness, but the p-value changing depending on the choice of whether to apply the X-perfect match or Y-perfect match test first is fascinating. The authors did discuss this very issue at several points in the manuscript. It is slightly unsettling to me that there isn't an exact p-value for a given set of data. This one point gives me a new perspective on statistics.

In full transparency, I don't believe I have to background to thoroughly review the appendix. I did read through it and did not notice any errors, but I couldn't confidently say there are not any small mathematical errors or any logical flaws in the proofs. Some sections were not easy to follow (my own shortcomings, the writing appeared sufficient for more of an expert to understand).

We greatly appreciate the reviewer’s time and effort tackling an appendix outside their comfort zone.

Reviewer #2 (Recommendations for the authors):

(1) In the numerical experiment session, the authors should include the null situation, i.e., the performance of the test when X and Y are independent. This helps assess the size of the test.

We have added a section on size to our results section, copied below:

“The permute-match test’s false positive rate depends on the process tested. The permute-match test is conservative – meaning that its false positive rate can fall below the significance level – because both the permutation test and perfect match test are conservative. As discussed elsewhere [29], the permutation test is conservative when α is not one of its possible p-values and when ties may occur between the original correlation and shuffled correlations. Checking for a perfect match is similarly conservative. The actual probability of a false-alarm perfect match event can vary depending on the process being tested. To see this consider the permute-match(X>Y) test in the setting where n = 3, where α = 0.05, and where rX,Y= 0 (independent X and Y). Note that in this case, obtaining a Y -perfect match (and thus p = 1/nn) is necessary and sufficient to detect dependence since α is too low for detection by either the permutation test or the p = 2/nn leg of the permute-match test. In the linear system of Fig 2, we observed among 5000 simulations a detection rate of 0.0148, significantly below the upper bound of 1/33 (Figure 2 - Source data 1; one-tailed exact binomial test, p < 10−20). Conversely, in the nonlinear system of Fig S2, this same event (Y -perfect match under n = 3 and rX,Y = 0) occurs with a detection rate of 0.0328, not significantly 9 below the upper bound of 1/33 (Figure S2 - Source data 1; one-tailed exact binomial test, p = 0.059). Thus, depending on the underlying process studied, the actual chance of a perfect match happening under independence may be near or significantly below the theoretical upper bound.”

(2) Some insights regarding the choice of rho should be provided. Especially, are there any examples that the classical test, such as the Pearson correlation or Granger causality test does not work?

We have redone the example of Appendix 4 with Pearson correlation, showing that Pearson correlation has substantially lower power than cross-map skill in this case (compare figures S2 and S3).

(3) Line 34 - 35, page 2: Correlations and causality should be separately considered. This sentence talks more about causality rather than correlation.

We appreciate the reviewer’s perspective and agree that correlation and causality are distinct.

We feel that pointing out the issue of spurious correlations is helpful to orient our readers, especially those from a broad scientific audience. In the text, we define “correlation” as a descriptive statistic (rather than normalized covariance), and later distinguish it from “dependence”, which has causal implications due to Reichenbach’s common cause principle. We believe this distinction provides a useful backdrop for practitioners who use statistical methods but are perhaps new to thinking deeply about statistical dependence.

(4) Please add some discussions on the situation that X_i depends on Y_{i - j} for some j > 0, which is associated with the setting of Granger causality test.

We have added the following to the discussion:

“No distributional assumptions are required, and the correlation function ρ can be completely arbitrary. For instance, ρ could include a lag to detect delayed dependence, or even evaluate the correlation strength at several lags and report the strongest among them [39].”

Version 2

Abstract

Researchers frequently analyze correlations between pairs of time series by determining whether an observed correlation is stronger than expected under the null hypothesis of independence. However, the time series are often nonstationary, with statistical properties that change over time, thereby making standard tests invalid. If sufficient replicates exist, a trial-swapping permutation test can be performed that handles nonstationarity by comparing within-replicate correlations to between-replicate correlations. Although largely assumption-free, this test is fundamentally limited by the number of replicates (n) because its minimum p-value is 1/n!. With n = 3, this minimum is 1/6, rendering thresholds like 0.05 unattainable. This limits its use considerably in animal experiments, where n may be as low as 3. We propose permute-match tests — modified permutation tests that can report lower p-values of 2/nn or 1/nn under strong evidence of dependence. Permute-match tests guarantee a false positive rate at or below the significance level when replicates are independent and identically distributed. The bound of 1/nn is not gratuitously conservative, since it cannot be further lowered without additional assumptions. We demonstrate our approach using synthetic data and apply it to an existing dataset with 3 independent groups of zebrafish, confirming the observation that zebrafish swim faster when directionally aligned.

Introduction

Scientists frequently look for correlations between variables to identify potentially important relationships or to support conceptual models. In disciplines with a focus on dynamics such as ecology and physiology, it is common to measure correlations between time series. Many important biological processes are nonstationary, meaning that their statistical properties (e.g., mean or variance) change systematically across time. Such processes range from the spread of an invasive species to a cell’s response to a new environmental stress.

Interpreting a correlation between a pair of nonstationary time series can be highly fraught because it is easy to obtain a seemingly impressive correlation between two time series that have no meaningful relationship [1]. For example, the sizes of any two exponentially-growing populations will be correlated over time due to a shared growth law, even if the populations lack any interactions or shared influences (e.g., bacterial cultures in separate tubes). To avoid overinterpreting spurious correlations, it helps to distinguish between the concepts of “correlation” and “dependence”. In time series research, the term “correlation” is often used procedurally [2, 3, 4]. Similarly, here we define a correlation function (ρ) to be any function that takes two time series and produces a statistic, although it is usually interpreted as a measure of similarity or relatedness.

In contrast to correlation, “dependence” is a hypothesis about the relationship between variables, and it has clearer scientific implications. Two events A and B are called dependent if P (A, B), the probability that they both occur, differs from P (A)P (B), the product of their individual probabilities. The formal definition of dependence extends this idea to continuous variables, where discrete probabilities are replaced by probability densities [5, 6]. Similarly, two temporal processes are dependent if the probability of observing any particular pair of trajectories differs from the product of the probabilities of the individual trajectories. The importance of dependence relationships can be attributed to Reichenbach’s common cause principle, which states that if two variables are dependent, then either they share a common causal influence, or one variable causally influences the other (possibly indirectly) [7, 8]. Thus, it is useful to first test whether an observed correlation indicates dependence before pursuing any mechanistic explanations.

There are a handful of systematic approaches to testing for dependence between nonstationary time series. However, most are fairly limited in their scope. First, when possible, a nonstationary time series can be transformed to become stationary, meaning that its statistical properties do not change systematically across time. This enables access to a wide arsenal of tests applicable to stationary data. Transformations include subtracting a trend, taking the derivative (more precisely, “differencing” between neighboring points), or choosing a stationary-seeming window of time [9, 10]. However, it is easy to see potential pathologies in each of these: Taking the derivative of an exponential curve just produces another exponential curve; subtracting a fitted linear trend from the random walk X(t) = X(t − 1) + ϵ(t) (where ϵ(t) is random noise) does not make it stationary [9]; similarly, searching for a stationary window of the same random walk process is futile since the variance of X(t) increases with each step. In a second approach, one may compare the observed correlation with correlations obtained when one time series is replaced by a synthetic replicate, where synthetic replicates are generated in a way that reflects the original nonstationary process [11]. However, these require a correct model of how the statistical properties of the process evolve over time. Lastly, there are tests within the econometrics literature that can provide evidence for dependence between random-walk-like nonstationary processes by detecting a property called cointegration, but cointegration only occurs when some linear combination of the time series is stationary [12].

Alternatively, a permutation-based approach is possible when there are multiple identically distributed and independent (iid) replicates (or trials; we use “replicates” and “trials” interchangeably). In this case, one may evaluate the significance of a within-replicate correlation by comparing it to between-replicate correlations. That is, if Xi and Yi are time series of variables X and Y from replicate i, the correlation of (Xi, Yi) may be compared to the correlation of (Xi, Yj≠i). If the two variables are dependent, within-replicate correlations may differ systematically from between-replicate correlations. This approach is sometimes called “inter-subject surrogates” [13, 14].

The inter-subject surrogate approach can be used to test for dependence in each trial separately [15]. In this case, a simple nonparametric test of dependence between, for instance, X1 and Y1 can be performed by computing the correlation ρ(X1, Yj) for j = 1, …, n and writing down a p-value as the proportion of these correlations that are greater than or equal to ρ(X1, Y1) [16, 17]. For this approach, at least n = 20 trials are needed if one wishes to possibly obtain a p-value of 0.05 or below. This procedure checks for dependence in trial 1 specifically; assessing dependence in a different trial via this method would involve an analogous test.

To reduce the required number of trials below 20, a straightforward strategy is to perform a single permutation test (Fig 1A) using the data from all replicates [18, 19]. As an example with n = 4 trials, the test procedure begins by computing the mean within-trial correlation:

ρorig=14(ρ(X1,Y1)+ρ(X2,Y2)+ρ(X3,Y3)+ρ(X4,Y4)). Next, as a null model, recompute the mean correlation after permuting the Y time series while holding the X time series in the original order. For instance, one such permutation might be: ρshuffle=14(ρ(X1,Y3)+ρ(X2,Y2)+ρ(X3,Y4)+ρ(X4,Y1)). A p-value can be calculated as the proportion of permutations (including the original ordering) that produce a mean correlation that is as strong as, or stronger than, ρorig (Fig 1A). One may then detect a correlation at the significance level α if p ≤ α. We emphasize that these permutations are obtained by swapping trials, not by swapping time points. This test has been used to detect correlations between time series in neuroscience and psychology settings [20, 18, 21]. It has also been used to detect correlations between variables measured in brain images, which can have similar nonstationarity challenges as time series [22]. A noteworthy advantage of this approach is that it is valid (meaning that the false positive rate is guaranteed to not exceed α) under very mild assumptions. Namely, the test is valid if the Xi trials are exchangeable with one another (i.e. all permutations of the sequence X1, …, Xn have the same joint probability distribution), or if the Yi trials are exchangeable (see Corollary 4 in Appendix 1).

The permutation test, perfect match concept, and permute-match tests.

(A) The permutation test. Circles indicate the average correlation without trial swapping (ρorig; pink) or with trial swapping (ρshuffle; grey). Here, N≥ is the number of average correlations (including ρorig) that are equal to or larger than the original correlation ρorig. It follows that pperm = N≥/n! because n! is the number of ways the trials can be shuffled. (B) Illustration of a Y-perfect match, represented as a table of correlations wherein each diagonal entry is the greatest in its own row. Each line with a “>” or “<” symbol denotes a required ordering relationship between the two numbers at either end of the line. Note that this example has a Y-perfect match, but not an X-perfect match. For an X-perfect match, each diagonal entry would need to be the greatest in its own column. (C) If an X- or Y-perfect match occurs, then the original correlation ρorig is always greater than any shuffled correlation ρshuffle. Top: If a Y-perfect match occurs, then each Yi gives the greatest correlation when it is paired with the Xi from the same trial. A shuffled correlation ρshuffle pairs Yis from some trials with Xjs from different trials, thereby reducing the correlation. Lines with symbols “>”, “<”, and “=” denote comparisons between terms. Bottom: If an X-perfect match occurs, a similar argument can be applied. (D) Since a perfect match ensures that the original correlation is greater than any shuffled correlation, a perfect match also ensures that pperm takes on its lowest possible value of 1/n!. (E) The simultaneous permute-match(X;Y) test. (F) The sequential permute-match(X▶Y) test; note that the permute-match(Y ▶X) test is defined analogously.

Yet, even the permutation test remains considerably limited by the number of trials. The lowest p-value that can be obtained with n replicates is 1/n! since there are n! possible permutations. For example, with n = 3, the lowest possible p-value is 1/3! ≈ 0.17 and with n = 4, the lowest is 1/4! ≈ 0.04. These minima apply even if the average within-trial correlation is much greater than the average between-trial correlation. While replicate counts of 4 and 5 can produce p-values below the conventional 0.05 threshold, these p-values remain modest. This limitation is particularly problematic when multiple hypothesis testing is required, as the modest p-values may no longer be significant after applying corrections such as the Bonferroni method.

The challenge of limited replicates is a reality in many experimental settings due to cost and feasibility concerns [23, 24, 25]. Factors outside the control of investigators further restrict the number of useful replicates. For example, a study of cerebral ischemia in pigs began with 6 animals, but lost 2 due to death and experimental mishap, so complete records exist for only 4 pigs [26]. Similarly, in an animal migration study, authors affixed GPS tags to 13 juvenile cuckoos, but only 5 of them produced GPS records suitable for the study due to heterogeneous behavior among cuckoos [27]. Finally, secondary analysis of public data from earlier studies can be a resource-efficient mode of research [28], although this approach is necessarily limited to the number of replicates within the existing dataset.

In this article, we show that permute-match tests, by adding additional steps to the permutation test, can achieve p-values as low as 1/nn. For n = 3 or 4 respectively, the lowest possible p-value is then ≈ 0.04 or 0.004, compared to 1/3! ≈ 0.17 or 1/4! ≈ 0.04. The permute-match tests only require that the replicates are iid. Thus, permute-match tests allow the data analyst to detect dependence with stronger confidence when there is sufficient evidence to do so. We illustrate this approach using simulations as well as a real world example involving recordings of zebrafish swimming behavior with only 3 replicates.

Results

The concept of an X-perfect or Y-perfect match

Central to our approach is the concept of an X-perfect match or Y-perfect match. To motivate the ideas, suppose that a group of n graduate students X = (X1, …, Xn) has been paired with a group of n advisors Y = (Y1, …, Yn) so that the ith student (Xi) is paired with the ith advisor (Yi). Moreover, let ρ(Xi, Yj) be a number that measures how well the ith student and the jth advisor get along. We say that the ith student is “happy” if they get along with their own advisor strictly better than they get along with any other advisor, meaning that ρ(Xi, Yi) >ρ(Xi, Yj) for all j≠ i. Similarly, the ith advisor is “happy” if ρ(Xi, Yi) >ρ(Xj, Yi) for all j≠ i. The arrangement of student-advisor pairs is X-perfect if all students are happy, and Y-perfect if all advisors are happy.

Analogously, if X = (X1, …, Xn) and Y = (Y1, …, Yn) are collections of time series with n trials each, and if ρ is some correlation function, we say that an X-perfect match has occurred if (and only if) for each i ∈ {1, …, n}, ρ(Xi,Yi)>ρ(Xi,Yj)for allj≠i. Similarly, a Y-perfect match has occurred if and only if for each i, ρ(Xi,Yi)>ρ(Xj,Yi)for allj≠i. See Fig 1B for an example of a Y-perfect (but not X-perfect) match.

Throughout the article, we use boldface X or Y to indicate a collection of trials, Xi or Yi to indicate an individual trial, and X or Y to refer to the variable in general.

A perfect match (either X-perfect or Y-perfect) provides especially strong evidence to support the hypothesis that X and Y are dependent. With a perfect match, the original correlation ρorig is strictly greater than any of the shuffled correlations, as illustrated in Fig 1C. Thus, a perfect match guarantees pperm = 1/n!, the lowest p-value achievable with the permutation test (Fig 1D). Furthermore, letting P denote probability, we can prove that P(X-perfect match)≤1/nn under the null hypothesis H0 wherein (1) the Yi trials are iid, (2) the Xi trials are iid, and (3) X and Y are independent (Lemma 8 in Appendix 2). In Proposition 14 (Appendix 2), we show that this upper bound of 1/nn cannot be reduced further without requiring additional assumptions.

The same bound applies to the Y-perfect match: under H0 we have P (Y −perfect match) ≤ 1/nn (Lemma 6 in Appendix 2) and thus (1) P(X-perfect match orY-perfect match)≤2/nn.

Permute-match tests: modified permutation tests of dependence

We now describe tests for dependence between nonstationary time series based on the ideas just laid out, which we will call permute-match tests. Suppose that an experiment with n replicates produces time series X = (X1, …, Xn) and Y = (Y1, …, Yn), and we want to test whether X and Y are dependent. For example, we may have video recordings of n animals, each in a separate but identically constructed enclosure; for the ith animal, we extract a time series of movement speed (Xi) and a time series of the distance from a light source (Yi). Using these data, we may wish to investigate whether the animals under study tend to move at different speeds depending on their distance from the light source. We introduce two tests: simultaneous permute-match and sequential permute-match. As we will see, the former can be more reproducible while the latter can be more powerful.

In the simultaneous permute-match test (Fig 1E), we check for both an X-perfect and a Y-perfect match. If either the X- or Y-perfect match occurs, we write down a p-value to be p = 2/nn as per equation 1. If neither X nor Y achieves a perfect match, then we fall back to the permutation test: p = pperm = N≥/n! where N≥ is the number of all n! possible permutations (including the original ordering) that produce an average correlation that is at least as large as the original ordering (Fig 1A).

In the sequential permute-match test, we begin by choosing a variable (either X or Y) for the first perfect match test. For example, suppose we choose X first (the “permute-match(X▶Y)” test). If an X-perfect match is observed, then p = 1/nn (Fig 1F). If X-perfect match is not observed, we then check for a Y-perfect match, and write p = 2/nn if a Y-perfect match is observed. If neither an X-perfect nor Y-perfect match occurs, then p = pperm. A permute-match(Y ▶X) test is defined analogously.

Under H0 (i.e. the Yi replicates are iid, the Xi replicates are iid, and X and Y are independent), all the above tests are valid, meaning that for any significance level α, we have P (p ≤ α) ≤ α. This fact is proven as Theorems 11-13 in Appendix 2.

The permute-match tests (Fig 1E-F) are “upgrades” to the permutation test (Fig 1A) because permute-match tests always report the same or a lower p-value (since 1/nn < 2/nn ≤ 1/n! for all integers n ≥ 2; see Lemma 10 in Appendix 2). In exchange for lower p-values, the permute-match tests have a slightly more restrictive data requirement than the permutation test. The permutation test is valid either if the Xi trials are exchangeable, or if the Yi trials are exchangeable (Corollary 4 in Appendix 1). The permute-match tests require that all Xi trials are iid and that Yi trials are iid. If trials are iid, then they are necessarily exchangeable. However, exchangeable trials are not always iid. For example, the first five cards drawn from a shuffled deck are exchangeable because all possible orderings of cards are equally likely, but they are not iid because if the first card is an ace of hearts, the second card cannot be the ace of hearts. In practice, biological experimental designs typically use replicates that are both exchangeable and iid, so the more restrictive requirement of the permute-match test is often inconsequential.

The different tests – permutation test, simultaneous permute-match test, and sequential permute-match tests – are valid tests of dependence given iid replicates (Theorems 11-13 in Appendix 2). In other words, they all guarantee false positive rates at or below the significance level. Two subtle aspects of the permute-match tests are worth mentioning here. First and crucially, the permutation test and perfect match events are nested. That is, an X- or Y-perfect match is possible only if pperm is already at its lowest possible value of 1/n! (Fig 1C-D). We can therefore combine the checks for perfect matches with the permutation test without inflating the false positive rate above the nominal level. This would not work if the tests were not nested, because then each test would separately contribute false positive results. The second subtlety refers specifically to the sequential permute-match tests. These tests use the lower p-value of 1/nn if the first perfect match occurs, but use the more conservative “Bonferroni-corrected” p-value of 2/nn if only the second match occurs (Fig 1F). This superficially resembles the practice of conducting additional tests merely because earlier tests did not result in a detection – a form of p-hacking that generally results in an invalid final p-value. However, our sequential permute-match procedure is actually valid because it combines binary events instead of continuous variables. See Appendix 3 for a detailed discussion of this point.

Permutation and permute-match test variants may differ in statistical power

Although the permutation test, simultaneous permute-match test, and sequential permute-match test are all valid, they vary in power – the probability of detecting true dependence. Depending on the number of replicates n and the desired significance level α, there are four regimes to consider (Fig 2A).

Statistical power of various permutation test variants as illustrated using a simple nonstationary system.

(A) Summary of test power as a function of the significance level α and the number of replicates n. (B, C) System equations and example dynamics. The processes X and Y are given by a linear trend with additive noise on a time grid t = 1, 2, …, 100. The noise terms ϵX,i(t) and ϵY,i(t) are drawn from a bivariate normal distribution with a mean of 0, variance of 1, and covariance of rX,Y. The chart shows an example pair of time series where X and Y are dependent (rX,Y = 0.3). (D) Statistical power for the permutation test and various permute-match tests as a function of the replicate number n, significance level α, and strength of dependence rX,Y. Power was estimated as the proportion of simulations in which dependence was detected, calculated from 5000 simulations at each value of rX,Y between rX,Y = 0 and 0.54 in steps of size 0.01. At rX,Y = 0, there is no dependence, so the curve at that point indicates the false positive rate rather than power. We chose the Pearson correlation coefficient as our correlation function ρ. See (A) for the color legend.

Regime 1: when α ≥ 1/n!, all tests have the same power. If the permutation test detects dependence, permute-match tests will too since they fall back to a permutation test in the “worst case” (Fig 2A, column 1). Conversely, if the permutation test did not detect dependence (i.e. pperm >α), then it follows that pperm > 1/n!, which implies that neither an X-perfect nor Y-perfect match occurred (Fig 1D), and so the permute-match tests cannot have detected dependence either.

Regime 2: when 1/n! >α ≥ 2/nn, all permute-match tests are tied for highest power (Fig 2A, column 2). In this regime, the permutation test has zero power, since its p-value is limited at 1/n!. Conversely, if either an X-perfect or a Y-perfect match occurs, then, the sequential permute-match tests as well as the simultaneous permute-match(X;Y) test will all detect dependence.

Regime 3: when 2/nn >α ≥ 1/nn, the simultaneous permute-match(X;Y) test will have no power as the desired p-value is below the lowest possible p-value achievable by the test. Only the permute-match(X▶Y) or (Y ▶X) tests now have power (Fig 2A, column 3). Note that permute-match(X▶Y) and permute-match(Y ▶X) test may differ in power, as can be seen in the upper left panel of Fig 2D. We do not currently know how to pre-diagnose which of these two tests will be most powerful based on a given dataset, so the choice of which test to run in this case is arbitrary.

Regime 4: when α < 1/nn, no test has any power even when a perfect match occurs.

Figure 2B-D illustrates these different scenarios using a simulated nonstationary system and some common significance cutoffs. We simulated time series of two variables (X and Y) from a small number of replicates (between 3 and 5). Here, the time series were produced from a pair of linear time trends with coupled additive noise (rX,Y determines the strength of coupling; Fig 2B). We chose the Pearson correlation coefficient as our correlation function ρ. The various sub-panels of Fig 2D illustrate the different aforementioned regimes.

The above example was chosen for ease of presentation: in fact, the example is so simple that it may be reasonably treated by just fitting and removing the linear trend. In Appendix 4, we show the overall higher statistical power of permute-match tests over the permutation test in a more complex example where the data-generating process is nonstationary (a logistic map with time-varying parameters) and where a nonlinear correlation function (cross-map skill) is preferable to (linear) Pearson correlation.

The permute-match test’s false positive rate depends on the process tested

The permute-match test is conservative – meaning that its false positive rate can fall below the significance level – because both the permutation test and perfect match test are conservative. As discussed elsewhere [29], the permutation test is conservative when α is not one of its possible p-values and when ties may occur between the original correlation and shuffled correlations. Checking for a perfect match is similarly conservative. The actual probability of a false-alarm perfect match event can vary depending on the process being tested.

To see this consider the permute-match(Y ▶X) test in the setting where n = 3, where α = 0.05, and where rX,Y = 0 (independent X and Y). Note that in this case, obtaining a Y-perfect match (and thus p = 1/nn) is necessary and sufficient to detect dependence since α is too low for detection by either the permutation test or the p = 2/nn leg of the permute-match test. In the linear system of Fig 2, we observed among 5000 simulations a detection rate of 0.0148, significantly below the upper bound of 1/33 (Figure 2 – Source data 1; one-tailed exact binomial test, p < 10−20). Conversely, in the nonlinear system of Fig S2, this same event (Y-perfect match under n = 3 and rX,Y = 0) occurs with a detection rate of 0.0328, not significantly below the upper bound of 1/33 (Figure S2 - Source data 1; one-tailed exact binomial test, p = 0.059). Thus, depending on the underlying process studied, the actual chance of a perfect match happening under independence may be near or significantly below the theoretical upper bound.

An example from animal behavior science

Living in groups is a common experience among animals. For example, at least half of fish species are thought to form groups at some stage of life [30, 31]. The properties of these groups can impart a variety of important fitness effects [32]. In groups of fish, coordinated swimming may facilitate foraging (e.g., by enabling fish to “communicate” the location of food to others) and predator escape (e.g., by confusing predators) [33, 31]. One study [31] observed, in videos of small groups of zebrafish (8 fish per group), that the fish swam faster when in a high-polarization state (i.e. directionally aligned) than when in a low-polarization state. A correlation between polarization and speed might indicate statistical dependence between the two variables, or it might just be a consequence of temporal trends (e.g., polarization and swimming speed both happen to decrease with time).

Using a publicly available dataset [34] containing fish trajectories from 3 replicate 10-minute videos of small groups of juvenile Danio rerio zebrafish (10 fish per group), we performed a permute-match test of dependence between polarization and fish swimming speed. As we will see, although swimming speed and the extent of polarization are potentially nonstationary, the permute-match test nevertheless detects a statistical dependence between them.

Directional polarization is quantified as the mean resultant length Ct of fish velocity [35]. We represent the swimming direction of each fish by a unit vector, and then define Ct as the length of the average unit vector (Fig. 4; Eq. 2). The mean resultant length is bounded between 0 (e.g., two opposite vectors) and 1 (e.g., two parallel vectors). To quantify speed, we took the “average individual speed”, which is a term we use to denote an average across individuals, not across time. More precisely, the average individual speed of a 10-fish group at time t is (s1,t + s2,t + · · · + s10,t)/10 where si,t is the speed of fish i at time t. Note that average individual speed is distinct from the speed of the group center, which has also been studied in Danio fish [32].

Neither the average individual speed nor the mean resultant length is obviously stationary (Fig 3A). By visual inspection, the average individual speed seems to decrease across time, and all time series are deemed nonstationary by a Kwiatkowski-Phillips-Schmidt-Shin test (p < 0.01), and so we may be on shaky ground to use methods that require stationarity. Additionally, since only 3 trials are available, the permutation test and the simultaneous permute-match test cannot detect dependence at the 0.05 level, so we perform a sequential permute-match test.

A sequential permute-match test detects dependence between average individual speed and alignment (as measured by mean resultant length) in small groups of zebrafish.

(A) Time series of average individual speed and mean resultant length for the three replicate videos. Black curves show original time series, and red curves show a 30-second moving average. Average individual speed appears to decrease with time in all trials. All 12 time series (2 variables × 3 trials × 2 smoothing conditions) were deemed non-stationary by a Kwiatkowski-Phillips-Schmidt-Shin (KPSS; [37]) test (p < 0.01). Note that the KPSS test seeks to reject a stationary null hypothesis in contrast to other common tests, whose null hypotheses are nonstationary. (B) Pearson correlation between alignment and average individual speed, both within trials and between trials, using all frames up to the length of the shortest speed and alignment time series (19308 frames; 10.06 minutes). Table entries are shaded by correlation. We observe a speed-perfect match (and an alignment-perfect match), and thus detect dependence with p = 1/33 ≈ 0.04. (C) A parametric test also detects dependence. As a parametric alternative, for each trial we averaged values of speed and mean resultant length over the complete time series (visualized in the scatter plot shown, where each point is one trial). The sample Pearson correlation of the time-averaged variables is 0.99996, and a one-tailed test of significance gives p ≈ 0.003. (D) Permute-match tests detected a significant correlation between speed and alignment more consistently than the parametric test. For a grid of lengths between 20 and 600 seconds we sampled 500 random segments of each length, each drawn from the first 600 seconds, and determined for each segment whether the parametric test and/or the two possible permute-match tests detected a significant (p ≤ 0.05) correlation. In the edge case of the maximum 600-second length, all 500 “random” segments were identical. We used the Pearson correlation as the correlation function ρ in the permute-match tests, and a one-tailed parametric test, reflecting the alternative hypothesis of correlation being positive (not merely nonzero).

The mean resultant length is a measure of alignment among angles.

Consider three unit vectors colored blue, yellow, and red. The length of net vector (dashed) divided by the number of vectors is defined as the mean resultant length. Compared to less-aligned angles (i), strongly aligned angles (ii) produce a longer net vector, and thus a greater mean resultant length.

We arbitrarily started with average individual speed and identified a “speed”-perfect match, indicating a significant correlation between average individual speed and mean resultant length (Fig 3B; p = 1/33 ≈ 0.04). Significance does not depend on the choice of which match is tested, since both a “speed”-perfect and a “mean resultant length”-perfect match are obtained. Note that although trials are independent of each other, all between-trial correlations from the full dataset are positive (off-diagonal elements in Fig 3B), illustrating the danger in applying naive data analysis methods to data that are autocorrelated or even nonstationary.

To try a parametric alternative, for each trial we time-averaged the average individual speed and mean resultant length over the full 10 minutes (Fig 3C; where each point is one trial). The sample Pearson correlation of the time-averaged variables is 0.99996, and a one-tailed test of significance (using the function stats.pearsonr from the Python package Scipy [36]) also detects dependence (p ≈ 0.003), consistent with the permute-match tests. This test has the null hypothesis that the two variables shown in Fig 3C (time-averaged average individual speed and time-averaged mean resultant length) are independent and Gaussian. We view this test as comparable to the permute-match tests because although it requires a distributional assumption, it can be valid even when the time series are nonstationary.

Since data are limited in many scientific applications, we explored whether the results of the sequential permute-match tests and the parametric Pearson correlation test were consistent across different time series lengths. We truncated time series to various lengths (from 20 seconds to 10 minutes), and checked whether each test reported a significant (p ≤ 0.05) correlation (Fig 3D). Both sequential permute-match tests reported a significant correlation at all lengths tested, whereas the parametric test was inconsistent, finding a significant correlation for about one third of lengths.

Discussion

Outside the context of time series, permutation tests have been celebrated because they require only minimal assumptions [38]. To satisfy the requirements of permutation tests, it is sufficient (see Corollary 4 in Appendix 1) to collect data from exchangeable replicates. Data in each trial can be autocorrelated (e.g., temporally or spatially) and nonstationary. No distributional assumptions are required, and the correlation function ρ can be completely arbitrary. For instance, ρ could include a lag to detect delayed dependence, or even evaluate the correlation strength at several lags and report the strongest among them [39]. Moreover, the ρ can be asymmetric so that ρ(X, Y)≠ ρ(Y, X) as occurs when correlations are based on techniques that use one time series to predict values of another (e.g., Appendix 4). Such freedom from assumptions stands in contrast to the situation in the analysis of single-replicate time series, where the practitioner may be forced to make assumptions that are difficult to verify. For example, statistical tests of correlation between two time series often require stationary data at a minimum [40]. Yet, it is difficult to verify that a single time series is stationary because stationarity is a property of an entire ensemble of time series [40], so checking for stationarity in a single time series is similar in spirit to checking whether a single data point is drawn from a Gaussian distribution.

However, applying a trial-swapping permutation test to check for significant correlations requires that a sufficient number of trials be available to permute amongst one another. When only small numbers of trials are available, the permutation test is severely limited by mathematics. The minimum p-value that can possibly be achieved by a permutation test is 1/n!, which will provide only modest evidence against a null hypothesis when n = 4 and is essentially useless when n = 3, regardless of how strong the true dependence may be. This limitation is particularly severe for researchers testing several related hypotheses, as this ideally requires a test that can produce p-values sufficiently low to maintain significance even after a multiple-testing correction.

Here we have described modified permutation tests – the sequential and simultaneous permute-match tests – that have access to lower p-values. In the sequential permute-match test, if the first variable does not display a perfect match, this test falls back to the simultaneous permute-match test. Thus, the sequential permute-match test is more powerful than the simultaneous test. However, the sequential test requires the arbitrary choice of checking for an X- or Y-perfect match first, making it potentially less reproducible between studies than the simultaneous test, which does not involve an arbitrary choice. If neither an X-nor Y-perfect match is obtained, both tests default to a permutation test. This step is valid because the tests are nested in the sense that a perfect match can only occur when the permutation p-value takes on its lowest possible value (Lemma 9 in Appendix 2).

The test appears computationally tractable for relevant sample sizes: Our implementation of the permute-match(X▶Y) procedure completed a single test of dependence in the setting of Fig 2 with an average runtime of 3 seconds when n = 10 on a 2023 14-inch MacBook Pro with an M2 Pro processor and 16 GB RAM (see Source data 1). For n > 10, a standard permutation test already can report a p-value below 3 × 10−8, so the perfect match test is likely unnecessary for typical applications.

It seems likely that the tests could be further modified to report an even lower p-value when an X- and Y-perfect match occur simultaneously, as in Fig 3B. However, we have not investigated further and this problem is left for future efforts.

Although here we focused on time series, the permute-match test can in principle be applied to other scenarios where iid replicates are available, but data within each replicate are dependent. One example is spatial data in brain imaging, where the permutation test has been performed using multiple scanned brains [22]. Beyond spatial processes, dependent data also appear in nucleotide sequences and natural language text. Overall, advances in methods that take advantage of multiple replicates may facilitate statistical testing without requiring assumptions that are difficult to verify.

Methods

Data preprocessing

We obtained fish trajectory data [34] from the web address at https://drive.google.com/drive/folders/1UmzlX-yJhzQ5KX5rGry8wZgXvcz6HefD. For the first trial we used the file at the path “10/1/trajectories_wo_gaps.npy” (where “10” indicates the number of fish and “1” indicates the first trial). For the second and third trials, the file path was the same except it began with “10/2/” and “10/3/” respectively. These files contain sequences of fish positions indexed by video frame, as well as constants such as the frame rate (32 frames per second) and approximate fish body length.

We estimated the velocity of each fish as v→i,t=p→i,t+1−p→i,t−12R where p→i,t is a 2-dimensional column vector specifying the position of the ith fish in units of body length at video frame t, v→i,t is a 2-dimensional column vector specifying the fish velocity in units of body length per second at video frame t, and R is the frame rate (32 frames per second). The 3 trials contained similar but slightly different numbers of frames. We truncated the v→i,t time series to the shortest among them (19308 frames ≈ 10.06 minutes) so they could be compared before downstream analyses.

A group’s average individual speed at each time was then given by 1N∑i=1N|v→i,t| where N is the number of fish. The mean resultant length Ct of a group at each time was given by the formula (2) Ct=1N|∑i=1Nv→i,t|v→i,t|| Velocity vectors v→i,t covering missing tracking data (fewer than 0.02% of such vectors in any trial) were excluded from calculations of average individual speed and mean resultant length. When calculating mean resultant length, fish with speed of exactly zero (whose direction of motion is undefined) were also excluded. Instances of zero speed were rare, occurring in at most 1 fish-frame in any trial. Average individual speed and mean resultant length were determined from the remaining fish (always at least 8 of 10), so no time point was dropped from any trial.

Statistical analysis

The permute-match test is described in Fig 1 and its validity is proved in Appendix 2. For the KPSS test, we used the function “statsmodels.tsa.stattools.kpss” in the Python package Statsmodels (ver. 0.10.1) [41]. We set the “regression” argument to “c” (which tests the null hypothesis of stationarity rather than trend-stationarity), and we set the “lags” argument to “auto” (which means that the number of lags used for testing stationarity is automatically chosen by a data-dependent method).

Artificial intelligence

ChatGPT 5.2 was used while constructing the proof of Lemma 15, particularly in the application of the “taking out what is known” property. Claude Opus 4.8 was used to review code, mathematical arguments, and article text for errors and unclear language. Claude Opus was also used to edit some parts of the manuscript for brevity. Corrections suggested by Claude Opus were reviewed by the authors and incorporated where appropriate.